Clines from PDE

### 2015-11-22 notes/pde-clines.Rmd
LD decay

```{r doc_setup, include=FALSE}
fig.dim <- 5
knitr::opts_chunk$set(fig.width=2*fig.dim,fig.height=fig.dim,fig.align='center')
library(ReacTran)
library(Matrix)
library(parallel)
source("tran.1D.R") # fix for log.A
```

@@

```{r plot_hap_probs, cache=TRUE, depends="solve_branching"}
hap.probs <- get_hap_probs( hap.soln, left=-.025, right=.025 )
image( hap.probs, zlim=c(0,1), legend=TRUE, yaxt='n', 
      main=paste(c("A:","B:"),paste(round(attr(hap.probs,"r"),3),collapse=" to ")) )
axis(2,at=(pretty(xgrid$x.mid)-min(xgrid$x.mid))/diff(range(xgrid$x.mid)),labels=pretty(xgrid$x.mid))
```

Distance to closest recombination in a heterozygote
---------------------------------------------------

A simpler statistic, that might approximate the correct quantity well,
is the distance along the chromosome from an $A$ allele until the closest recombination
that occurred in an $AB$ heterozygote.
Let $H$ be this distance;
then, 
$$\begin{aligned}
    u(x,\tau,r)=\P^{(x,\tau)}\{H>r\}
\end{aligned}$$
solves the following PDE:
$$\begin{aligned}
    u(x,0,r) &= 1  \\
    \partial_t u(x,t,r)
        &= 
        \grad \log(p(x)) \cdot \grad u(x,t,r) + \frac{\sigma^2}{2} \Delta u(x,t,r) - r (1-p(x)) u(x,t,r) .
\end{aligned}$$

>>>>

```{r doc_setup, include=FALSE}
fig.dim <- 5
knitr::opts_chunk$set(fig.width=2*fig.dim,fig.height=fig.dim,fig.align='center')
library(ReacTran)
library(Matrix)
source("tran.1D.R") # fix for log.A
```

@@

Here are haplotype probabilities centered on the selected site:
```{r plot_hap_probs, cache=TRUE, depends="solve_branching"}
for (haplen in c(0.02, 0.26, 0.46)) {
    hap.probs <- get_hap_probs( hap.soln, left=-haplen/2, right=haplen/2 )
    image( hap.probs, zlim=c(0,1), legend=TRUE, yaxt='n', 
          main=paste(c("A:","B:"),paste(round(attr(hap.probs,"r"),3),collapse=" to ")) )
    axis(2,at=(pretty(xgrid$x.mid)-min(xgrid$x.mid))/diff(range(xgrid$x.mid)),labels=pretty(xgrid$x.mid))
}
```

And, here are haplotype probabilities for a fairly short haplotype moving away from the selected site:
```{r plot_hap_probs_locs, cache=TRUE, depends="solve_branching"}
haplen <- 0.04
for (haploc in c(0.01, 0.09, 0.21, 0.31)) {
    hap.probs <- get_hap_probs( hap.soln, left=haploc, right=haploc+haplen )
    image( hap.probs, zlim=c(0,1), legend=TRUE, yaxt='n', 
          main=paste(c("A:","B:"),paste(round(attr(hap.probs,"r"),3),collapse=" to ")) )
    axis(2,at=(pretty(xgrid$x.mid)-min(xgrid$x.mid))/diff(range(xgrid$x.mid)),labels=pretty(xgrid$x.mid))
    locator(1)
}
```

Here's all the solution traces, i.e., how probabilities of inheriting an $A$ haplotype
of all lengths, positions, and spatial locations changes with time
```{r plot_hap_everything, cache=TRUE, depends="solve_branching"}
matplot(hap.soln[,1],hap.soln[,-1],type='l',xlab='time',ylab='probability')
```

Here's how these look across space, at a fixed time,
with line type showing haplotype length
and color showing distance to the selected site.
```{r plot_hap_clines, cache=TRUE, depends="solve_branching"}
plot.t <- 3
hap.clines <- hap.soln[which(hap.soln[,1]>=plot.t)[1],-1]
dim(hap.clines) <- attr(hap.soln,"soln.dims")
r.df <- data.frame(attr(hap.soln,"r"))
r.df$length <- r.df$right-r.df$left
r.df$dist <- ifelse(r.df$right*r.df$left>0,pmin(abs(r.df$right),abs(r.df$left)),0)
r.df$col <- as.numeric(cut(r.df$dist,6))
r.df$lty <- as.numeric(cut(r.df$length,5))
matplot(xgrid$x.mid,hap.clines[1:xgrid$N,],type='l',main=paste("A: t=",plot.t),ylim=c(0,1),xlab="space",col=r.df$col,lty=r.df$lty,ylab="probability")
legend("bottomleft",lty=unique(r.df$lty),legend=paste("length=",round(tapply(r.df$length,r.df$lty,mean),2)))
matplot(xgrid$x.mid,hap.clines[xgrid$N+1:xgrid$N,],type='l',main=paste("B: t=",plot.t),ylim=c(0,1),xlab="space",col=r.df$col,lty=r.df$lty,ylab="probability")
legend("topright",lty=1,col=unique(r.df$col),legend=paste("dist=",round(tapply(r.df$dist,r.df$col,mean),2)))
```

And, here's a heatmap of haplotype length across space and time:
shown is the longest haplotype length that has probability of at least 50% of coming entirely from $A$:
```{r plot_hap_lens, cache=TRUE, depends=c("solve_branching","plot_hap_clines")}
xfac <- rep(1:xgrid$N,nrow(r.df))
ABfac <- rep( rep(c("A","B"),each=xgrid$N), nrow(r.df) )
hap.lens <- lapply( c(A="A",B="B"), function (allele) {
    do.call( rbind, lapply( 1:nrow(hap.soln), function (k) {
            tapply( hap.soln[k,1+which(ABfac==allele)], xfac, function (xx) {
                   if(any(xx>0.5)) { max(r.df$length[xx>0.5]) } else { 0.0 }
                } )
        } ) )
    } )
filled.contour(hap.soln[,1],xgrid$x.mid,hap.lens$A,xlab="time",ylab="space",zlim=c(0,max(hap.lens$B)),main="A-linked A haplotype length")
filled.contour(hap.soln[,1],xgrid$x.mid,hap.lens$B,xlab="time",ylab="space",zlim=c(0,max(hap.lens$B)),main="B-linked A haplotype length")
```

Linkage
=======

The way the discretization for haplotypes is implemented
effectively only keeps track of a finite number of markers.
To do two-site ancestral LD, then,
we just need to use the above code at only two sites on the genome.

First, we compute per-allele solutions:
```{r solve_LD, cache=TRUE}
s <- 0.1
tt <- seq(0,100,length.out=21)
xgrid <- setup.grid.1D(x.up=-20.5, x.down=20.5, N=41)
rr <- seq(0,0.5,length.out=20)[-1]
fwds.soln <- forwards_pde(s=s,times=tt,grid=xgrid,yinit=ifelse(xgrid$x.mid<0,1,0))
LD.solns <- lapply( rr, function (r) forwards_backwards_haplotypes(s=s, times=tt, xgrid=xgrid, rgrid=list(x.mid=c(0,r),N=2), fwds.soln=fwds.soln) )
```
... but then we must extract linkage from these.
Below we report for recombination distance $r$,
the probability that two markers at distance $r$ 
sampled in one individual at that location and time
(with one end fixed to the selected site)
are both of the same ancestral type.
If the local proportion of type $A$ is $p$ and the probability when linked to $A$ is $u_A$,
we want $p u_A(r_0,r_0+r) + (1-p) u_B(r_0,r_0+r) - (p u_A(r_0) + (1-p) u_A(r_0))(p u_A(r_0+r) + (1-p) u_A(r_0+r))$:
(**is that right?**)
```{r compute_LD, cache=TRUE, depends="solve_LD"}
pfun <- function (t,x) { 
        fp <- function (x,t) { approx(xgrid$x.mid,cline_interp(t,soln=fwds.soln),xout=x)$y }
        ans <- numeric(length(x))
        for (tval in unique(t)) {
            dothese <- (rep_len(t,length(x))==tval)
            ans[dothese] <- fp(x=x[dothese],t=tval)
        }
        return(ans)
}
ABfac <- rep( rep(c("A","B"),each=xgrid$N), 3 )  # three combinations of two recomb points
rfac <- rep( c("11","12","22"), each=xgrid$N*2 )
LDs <- lapply( LD.solns, function (lds) {
            # returns a time x space matrix
            pp <- outer( lds[,1], xgrid$x.mid, pfun )
            return( 
                   pp * lds[,1+which(ABfac=="A"&rfac=="12")] + (1-pp) * lds[,1+which(ABfac=="B"&rfac=="12")] 
                   - ( pp * lds[,1+which(ABfac=="A"&rfac=="11")] + (1-pp) * lds[,1+which(ABfac=="B"&rfac=="11")] )
                       * ( pp * lds[,1+which(ABfac=="A"&rfac=="22")] + (1-pp) * lds[,1+which(ABfac=="B"&rfac=="22")] )
               )
    } )
plot.x <- xgrid$x.mid[which.min(abs(xgrid$x.mid))]
matplot( tt, sapply( LDs, function (x) { x[,match(plot.x,xgrid$x.mid)] } ), type='l', ylab="ancestral LD", xlab="time" )
plot.t <- tt[which.min(abs(tt-20))]
matplot( xgrid$x.mid, sapply( LDs, function (x) { x[match(plot.t,tt),] } ), type='l', ylab="ancestral LD", xlab="space" )
matplot( rr, t(sapply( LDs, function (x) { x[match(plot.t,tt),] } )), type='l', ylab="ancestral LD", xlab="recomb dist", main=paste("t=",round(plot.t,2)), col=rainbow(xgrid$N), lty=1 )
legend("topright",lty=1,col=rainbow(xgrid$N)[floor(seq(1,xgrid$N,length.out=6))],legend=paste("x=",round(xgrid$x.mid[floor(seq(1,xgrid$N,length.out=6))],2)))
```
Hm, the right/left spatial positions should look the same, right?  Is something off?

Distance to closest recombination in a heterozygote
===================================================

A simpler statistic, that might approximate haplotype statistics well,
is the distance along the chromosome from an $A$ allele until the closest recombination
that occurred in an $AB$ heterozygote.
Let $H$ be this distance;
then, 
$$\begin{aligned}
    u(x,\tau,r)=\P^{(x,\tau)}\{H>r\}
\end{aligned}$$
solves the following PDE:
$$\begin{aligned}
    u(x,0,r) &= 1  \\
    \partial_t u(x,t,r)
        &= 
        \grad \log(p(x)) \cdot \grad u(x,t,r) + \frac{\sigma^2}{2} \Delta u(x,t,r) - r (1-p(x)) u(x,t,r) .
\end{aligned}$$
